Updated Criteria for the Diagnostic Procedure for Parkinson’s Disease Dementia on Level I and their Validity in Deep Brain Stimulation Cohort

### 2025-05-07 _manuscript/manuscript.qmd

Despite its clinical relevance, the diagnosis of PDD remains complex. *The original **International Parkinson and Movement Disorder Society (MDS)** diagnostic criteria [@dubois2007], first formalized in 2006*^[**JM: I ain't sure it was the first time PDD criteria were formalised; e.g., Aarsland et al. - dx.doi.org/10.1001/archneur.60.3.387 - used DSM criteria for PDD already in 2003. Consequently I added the specification that these were MDS crtieria.**], were heavily influenced by frameworks established primarily for Alzheimer’s disease (AD). Although these criteria provided a valuable initial foundation, they lacked the specificity required to capture the distinct pathophysiological and cognitive features of PD-related dementia [@emre2007; @yamashita2023].

*One notable feature of the original criteria was the provision of an algorithm that allowed for flexibility in test selection [@dubois2007]*^[**JM: This came out of nowhere according to me. I think it would make more sense to introduce level I vs II first then introduce the algortihm for level I (i.e., Dubois's Table 2) and focus on it.**]. Specifically, clinicians could choose between months reversed or seven backwards for attention assessment, lexical fluency or clock drawing for executive function evaluation. Moreover, the critaria included MMSE pentagons for evaluation of visuospatial ability, and three-word recall for memory assessment. Agreement across different criteria allows for the parallel computation of inter-rater reliability , which, in turn, facilitates the calculation of construct validity [@conway1995] further strengthening the diagnostic framework for PDD.

*Subsequent steps have introduced the concept of Parkinson’s disease mild cognitive impairment (PD-MCI), refining the understanding of cognitive dysfunction in PD [@litvan2012]. The PD-MCI criteria have propelled progress in diverse areas, enabling improvements in clinical characterization, identification of genetic correlates, therapeutic interventions, clinical trial design, and the assessment of progression risk to PDD [@hoogland2017; @hoogland2019; @aarsland2021].*^[**JM: I do not believe this section is needed for the article. You could keep it for yourself and use it in your dissertion but discussion history of MCI, albeit briefly, is out of scope of this manuscript in my humble correct opinion.**]

Currently, efforts are focused on refining the PDD diagnostic framework to improve its consistency and applicability in both research and clinical contexts [@kulisevsky2024]. *Our prior investigations have underscored the importance of employing rigorous psychometric methodologies to differentiate PD-MCI from PDD [@bezdicek2016; @deuschl2006].*^[**JM: The citations do not seem to match the statement - Bezdicek (2016) is about a processing speed theory of cognitive impairment in healthy older adults, Deuschl (2006) is an RCT of DBS. I'd drop this sentence altogether.**]

*According to a recent metanalysis, the epidemiological estimates of PDD prevalence among individuals with PD vary widely, ranging from 14% to 55%^[**JM: It makes sense to report some point estimate as well.**], depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients` sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline [@oh2016; @rana2012].*^[**JM: Good, I'd push this paragraph higher.**]

The present study evaluates the diagnostic concordance between the original Level I PDD criteria, as established by the MDS Task Force [@emre2007; @dubois2007] and criteria inspired by the recent call for change [@kulisevsky2024] within a cohort of PD patients *selected for DBS*^[**JM: Not true, the cohort is of patients being considered for DBS, not selected for DBS.**]. Furthermore, both sets of criteria are compared to PDD diagnosed on Level II. The study aims to address following research objectives (RO): 1) to estimate the prevalence of PDD among patients considered for DBS (RO1), 2) to assess variability in PDD diagnosis depending on the diagnostic criteria applied (RO2), 3) to evaluate the diagnostic concordance between different sets of PDD criteria (RO3), and 4) to identify diagnostic components that contribute to variability in PDD classification across criteria (RO4). By addressing these objectives, this study seeks to validate the revised PDD criteria and evaluate their relevance in the context of DBS eligibility, thereby contributing to the refinement of cognitive assessment protocols in PD.

Suggestion: (MK) **(RO1) To estimate the prevalence of Parkinson’s disease dementia (PDD) and evaluate the diagnostic variability and concordance across different operationalizations of probable PDD. (RO2) To identify specific diagnostic components that contribute to variability in PDD classification across the applied criteria.**^[**JM:Nice! Let's state it this way.**]

@@

## Diagnostic algorithms for probable Parkinson's Disease Dementia^[**JM: Add a section with some logic (Cartezian product) to Appendix as well.**]

@@

Next, a set of two class cross-tabulations with associated statistics was computed for each pair of operationalizations via the `confusionMatrix()` function from the R package *caret* [@kuhn2008]. For each pair of operationalizations, the analysis was repeated twice such that each variables of the pair served once as the reference and once as the predictor. Following measures were used to evaluate pairwise concordance between different operationalizations of PDD criteria: 1) Cohen's $\kappa$ with its 95% confidence interval (CI) computed via the `cohen.kappa()` function from the R package *psych* [@revelle2024]; 2) Accuracy (i.e., the proportion of correct predictions, both true positives and true negatives, among the total number of cases) with its 95% CI; 3) Sensitivity/Recall (i.e., the proportion of true positives); and 4) Specificity (i.e., the proportion of true negatives).[^4]

[^4]: Unlike Cohen's $\kappa$, Accuracy, Sensitivity and Specificity are not symmetrical, i.e., their value depend on which variable is considered reference and which is considered predictor. Consequently, we report these values twice for each pair of operationalizations. Note that the Sensitivity of a reference/predictor pair corresponds to the Positive Predictive Value if their roles were reversed. The same relationship holds true between the Specificity and the Negative Predictive Value.

>>>>

Despite its clinical relevance, the diagnosis of PDD remains complex. The original International Parkinson and Movement Disorder Society (MDS) diagnostic criteria [@dubois2007], first formalized in 2007, were heavily influenced by frameworks established primarily for Alzheimer’s disease (AD). Although these criteria provided a valuable initial foundation, they lacked the specificity required to capture the distinct pathophysiological and cognitive features of PD-related dementia [@emre2007; @yamashita2023].

To enhance diagnostic accuracy, the MDS introduced a two-level system for Parkinson’s disease dementia. Level I criteria are designed for clinical use, relying on brief cognitive assessments and clinical judgment. Level II criteria involve comprehensive neuropsychological testing across multiple domains, providing greater diagnostic detail for research or specialized settings [@emre2007].

One notable feature of the original Level I criteria was the provision of an algorithm that allowed for flexibility in test selection [@dubois2007]. Specifically, clinicians could choose between months reversed or seven backwards for attention assessment, lexical fluency or clock drawing for executive function evaluation. Moreover, the critaria included MMSE pentagons for evaluation of visuospatial ability, and three-word recall for memory assessment. Agreement across different criteria allows for the parallel computation of inter-rater reliability , which, in turn, facilitates the calculation of construct validity [@conway1995] further strengthening the diagnostic framework for PDD.

According to a recent metanalysis, the epidemiological estimates of PDD prevalence among individuals with PD vary widely, ranging from 14% to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients` sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline [@oh2016; @rana2012].

Currently, efforts are focused on refining the PDD diagnostic framework to improve its consistency and applicability in both research and clinical contexts [@kulisevsky2024].

The present study evaluates the diagnostic concordance between the original Level I PDD criteria, as established by the MDS Task Force [@emre2007; @dubois2007] and criteria inspired by the recent call for change [@kulisevsky2024] within a cohort of PD patients considered for DBS. Furthermore, both sets of criteria are compared to PDD diagnosed on Level II. The study aims to address following research objectives (RO): (RO1) To estimate the prevalence of Parkinson’s disease dementia (PDD) and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components that contribute to variability in PDD classification across the applied criteria.

@@

## Diagnostic algorithms for probable Parkinson's Disease Dementia

@@

Next, a set of two class cross-tabulations with associated statistics was computed for each pair of operationalizations via the `confusionMatrix()` function from the R package *caret* [@kuhn2008]. For each pair of operationalizations, the analysis was repeated twice such that each variables of the pair served once as the reference and once as the predictor. Following measures were used to evaluate pairwise concordance between different operationalizations of PDD criteria: 1) Cohen's $\kappa$ with its 95% confidence interval (CI) computed via the `cohen.kappa()` function from the R package *psych* [@revelle2024]; 2) Accuracy (i.e., the proportion of correct predictions, both true positives and true negatives, among the total number of cases) with its 95% CI; 3) Sensitivity/Recall (i.e., the proportion of true positives); and 4) Specificity (i.e., the proportion of true negatives).[^10]

[^10]: Unlike Cohen's $\kappa$, Accuracy, Sensitivity and Specificity are not symmetrical, i.e., their value depend on which variable is considered reference and which is considered predictor. Consequently, we report these values twice for each pair of operationalizations. Note that the Sensitivity of a reference/predictor pair corresponds to the Positive Predictive Value if their roles were reversed. The same relationship holds true between the Specificity and the Negative Predictive Value.
